WHOOMP! (There It Is) \\ Rapid Bayesian position reconstruction for gravitational\nobreakdashes-wave transients

### 2014-05-28 ms.tex
Describe BAYESTAR likelihood

First, we put the phase and time on the same footing by measuring the time in units of $1 / \sqrt{\overline{\omega^2}}$ with a change of variables from $\tau$ to $\gamma_\tau = \sqrt{\overline{\omega^2}} \tau$:
%
\begin{equation}
    \mathcal{I}' = \bordermatrix{
        ~ & \rho_i & \gamma_i & \gamma_{\tau,i} \cr
        \rho_i & 1 & 0 & 0 \cr
        \gamma_i & 0 & {\rho_i}^2 & -{\rho_i}^2\frac{{\overline{\omega}}_i}{\sqrt{{\overline{\omega^2}}_i}} \cr
        \gamma_{\tau,i} & 0 & -{\rho_i}^2\frac{{\overline{\omega}}_i}{\sqrt{{\overline{\omega^2}}_i}} & {\rho_i}^2
    }.
\end{equation}
%
The second change of variables, from $\gamma$ and $\gamma_\tau$ to $\gamma_\pm = \frac{1}{\sqrt{2}}(\gamma \pm \gamma_\tau)$, diagonalizes the Fisher matrix:
%
\begin{equation}
    \mathcal{I}'' = \bordermatrix{
        ~ & \rho_i & \gamma_{+,i} & \gamma_{-,i} \cr
        \rho_i & 1 & 0 & 0 \cr
        \gamma_{+,i} & 0 & \left(1 - \frac{\overline{\omega}_i}{\sqrt{\overline{\omega^2}_i}}\right){\rho_i}^2 & 0 \cr
        \gamma_{-,i} & 0 & 0 & \left(1 + \frac{\overline{\omega}_i}{\sqrt{\overline{\omega^2}_i}}\right){\rho_i}^2
    }.
\end{equation}
%
Thus, in the appopriate time units, the \textit{sum and difference} of the phase and time of the signal are measured independently.

\subsection{Noise model for point parameters}

The likelihood in \Eref{eq:gaussian-likelihood-spa} is the quantity that is maximized by the matched\nobreakdashes-filter estimator. However, just because we have the covariance matrix defined by \Eref{eq:covariance-matrix} does not mean that the distribution of estimator errors is multivariate Gaussian. Nonetheless, it is straightforward to qualitatively describe the distribution.

If we were to suppress the time parameter $\tau$, then we could reduce the likelihood to the form
%
\begin{equation}
    \mathcal{L}(z; \rho, \gamma) \propto \exp \left[ -\frac{1}{2} \left|z - e^{i \gamma} \rho \right|^2 \right]
\end{equation}
%
for a complex scalar measurement $z$. This likelihood also describes the distribution of the length $\rho$ and angle $\gamma$ of a vector whose Cartesian components are perturbed with normally distributed errors. For this problem, $\rho$ and $\gamma$ are uncorrelated, $\rho^2$ being noncentrally $\chi^2$\nobreakdashes-disributed with two degrees of freedom and $\gamma$ being von~Mises\nobreakdashes-distributed.%
%
\marginpar{Picture here?}

\marginpar{What about if we suppress $\gamma$ or $\rho$? Do those cases teach us anything?}%
%
We know that the matched\nobreakdashes-filter estimator's distribution, given a particular true value of $\rho$, $\gamma$, and $\tau$, must go to zero at $\hat{\rho}=0$ and must be periodic in $\gamma$ but not $\tau$. From just these properties, the estimator's distribution should be an approximate Gaussian, but wrapped diagonally around a cylinder where the radius represents $\hat{\rho}$, the angle around the cylinder represents $\hat{\gamma}$, and the distance up or down the cylinder represents $\hat{\tau}$.

There are many ways to realize a distribution function with these properties, but they tend to be difficult to work with.
%
\marginpar{Give some examples.}%
%
An alternative is to just discard the one periodic parameter, $\gamma$, entirely. We do not know the exact distribution of the remaining parameters $\rho$ and $\tau$, but we are less liable to make egregious errors by assuming that they are Gaussian than if we assumed that the estimation errors in $\gamma$ were Gaussian. Projecting out $\gamma$ gives us the covariance matrix
%
\begin{equation}
    \cov \left(
        \begin{array}{c}
            \hat{\rho} - \rho \\
            \hat{\tau} - \tau
        \end{array}
    \right) =
    \left(
        \begin{array}{cc}
            1 & 0 \\
            0 & \rho^2/{\omega_\mathrm{rms}}^2
        \end{array}
    \right).
\end{equation}

A wrinkle is that this covariance matrix still depends on the true, unknown value of $\rho$. If we make the further leap of evaluating the \ac{CRLB} \emph{at the value of the estimators}, we can then write the distribution of the estimator error as
%
\begin{eqnarray}
    \hat{\rho} - \rho \sim \mathcal{N}\, [0, 1] \\
    \hat{\tau} - \tau \sim \mathcal{N} \left[0, \hat{\rho}^2 / {\omega_\mathrm{rms}}^2\right],
\end{eqnarray}
%
valid at moderate to high \ac{SNR}.

\marginpar{It's difficult to follow from here where the timing uncertainty is used in the following section.}

>>>>

First, we put the phase and time on the same footing by measuring the time in units of $1 / \sqrt{\overline{\omega^2}}$ with a change of variables from $\tau$ to $\gamma_\tau = \sqrt{\overline{\omega^2}} \tau$:
%
\begin{equation}
    \mathcal{I}' = \bordermatrix{
        ~ & \rho_i & \gamma_i & \gamma_{\tau,i} \cr
        \rho_i & 1 & 0 & 0 \cr
        \gamma_i & 0 & {\rho_i}^2 & -{\rho_i}^2\frac{{\overline{\omega}}_i}{\sqrt{{\overline{\omega^2}}_i}} \cr
        \gamma_{\tau,i} & 0 & -{\rho_i}^2\frac{{\overline{\omega}}_i}{\sqrt{{\overline{\omega^2}}_i}} & {\rho_i}^2
    }.
\end{equation}
%
The second change of variables, from $\gamma$ and $\gamma_\tau$ to $\gamma_\pm = \frac{1}{\sqrt{2}}(\gamma \pm \gamma_\tau)$, diagonalizes the Fisher matrix:
%
\begin{equation}\label{eq:fisher-matrix-extrinsic-one-detector}
    \mathcal{I}'' = \bordermatrix{
        ~ & \rho_i & \gamma_{+,i} & \gamma_{-,i} \cr
        \rho_i & 1 & 0 & 0 \cr
        \gamma_{+,i} & 0 & \left(1 - \frac{\overline{\omega}_i}{\sqrt{\overline{\omega^2}_i}}\right){\rho_i}^2 & 0 \cr
        \gamma_{-,i} & 0 & 0 & \left(1 + \frac{\overline{\omega}_i}{\sqrt{\overline{\omega^2}_i}}\right){\rho_i}^2
    }.
\end{equation}
%
Thus, in the appopriate time units, the \textit{sum and difference} of the phase and time of the signal are measured independently.

\section{The \ac{BAYESTAR} likelihood}

Although we may not be able to compute the detection candidate likelihood, \Eref{eq:detection-candidate-likelihood}, directly, we can make an educated guess that has many properties in common with it. Any valid approximate likelihood must have the same Fisher matrix \Eref{eq:fisher-matrix-extrinsic-one-detector}. It must also have the same limiting behavior: it should be periodic in the phase error $\tilde{\gamma}_i$ and go to zero as $\tilde{\tau}_i \rightarrow \pm \infty$, $\hat{\rho}_i \rightarrow 0$, or $\hat{\rho}_i \rightarrow \infty$. Additionally, when $\tilde{\tau}_i = 0$, the distribution of ${\hat{\rho}_i}^2$ should reduce to a noncentral $\chi^2$ distribution with two degrees of freedom, centered about ${\rho_i}^2$, because the complex matched filter time series $y_i(t)$ is Gaussian.

These conditions could be satisfied by realizing a multivariate Gaussian distribution with covariance matrix $\Sigma = \mathcal{I}^\intercal$, and then replacing individual terms of the form $-\tilde{\theta}^2/2$ with $\cos{\tilde{\theta}}$.

\marginpar{Actually, this does not give exactly the same Fisher matrix. Explain.}%
%
Another way is to plug the signal model from \Eref{eq:signal-model} \emph{evaluated at the ML parameter estimates} into the single-detector likelihood in \Eref{eq:gaussian-likelihood-spa}:
%
\begin{eqnarray*}
	\fl p\left(\hat{\boldsymbol\theta}_i \middle| \boldsymbol\theta \right)
	&:=&
	p\left(x_i(\omega) = h_i(\omega; \hat{\boldsymbol\theta}_i)
		\middle| \boldsymbol\theta \right)
	\\\fl
	&\propto& \exp \left[
		- \frac{1}{2} \int_0^\infty \frac{\left|
			\frac{\hat{\rho}_i}{\sigma_i(\hat{\boldsymbol\theta}_\mathrm{in})} e^{i (\hat\gamma_i - \omega \hat\tau_i)} h(\omega; \hat{\boldsymbol\theta}_\mathrm{in})
			- \frac{\rho_i}{\sigma_i(\boldsymbol\theta_\mathrm{in})} e^{i (\gamma_i - \omega \tau_i)} h(\omega; \boldsymbol\theta_\mathrm{in})
		\right|^2}{S_i(\omega)} \, d\omega
	\right].
\end{eqnarray*}
%
If we assume that $\hat{\boldsymbol\theta}_\mathrm{in} = \boldsymbol\theta_\mathrm{in}$, then this reduces to
%
\begin{equation}
	p\left(\hat\rho_i, \hat\gamma_i, \hat\tau_i
		\middle| \rho_i, \gamma_i, \tau_i \right) \propto
	\exp \left[ - \frac{1}{2}{\hat\rho_i}^2 - \frac{1}{2}{\rho_i}^2
		+ \hat\rho_i \rho_i \Re \left\{ e^{i \tilde{\gamma}_i} a^*(\tilde{\tau}_i)
		\right\}
	\right]
\end{equation}
%
with $\tilde{\gamma}_i = \hat\gamma_i - \gamma_i$, $\tilde{\tau}_i = \hat\tau_i - \tau_i$, and the template's autocorrelation function $a(t; \boldsymbol\theta_\mathrm{in})$ defined as
%
\begin{equation}
	a(t; \boldsymbol\theta_\mathrm{in}) := \frac{1}{\sigma_i(\boldsymbol\theta_\mathrm{in})} \int_0^\infty \frac{\left| h(\omega; \hat{\boldsymbol\theta}_\mathrm{in})\right|^2}{S_i(\omega)} e^{i \omega t} \,d\omega.
\end{equation}

To assemble the joint likelihood for the whole network, we just form the product from the individual detectors:
%
\begin{equation}
	\fl p\left(\left\{\hat\rho_i, \hat\gamma_i, \hat\tau_i\right\}_i
		\middle| \left\{\rho_i, \gamma_i, \tau_i\right\}_i \right) \propto
	\exp \left[ - \frac{1}{2} \sum_i {\hat\rho_i}^2 - \frac{1}{2} \sum_i {\rho_i}^2
		+ \sum_i \hat\rho_i \rho_i \Re \left\{ e^{i \tilde{\gamma}_i} a^*(\tilde{\tau}_i)
		\right\}
	\right].
\end{equation}
%
To marginalize over the coalescence phase, we can insert an arbitrary nuisance phase $\eta$ by making the replacement $\tilde{\gamma}_i \rightarrow \tilde{\gamma}_i + \eta$. Then integrating over $\eta$ from $0$ to $2\pi$ and suppressing normalization factors, we get the \ac{BAYESTAR} likelihood
%
\begin{equation}
	\mathcal{L}_\mathrm{BAYESTAR} = \exp \left[ - \frac{1}{2} \sum_i {\rho_i}^2 \right] I_0 \left[
		\left| \sum_i \hat\rho_i \rho_i e^{i \tilde{\gamma}_i} a^*(\tilde{\tau}_i)
		\right|
	\right],
\end{equation}
%
where $I_0$ is a modified Bessel function of the first kind.
